---
title: "Initial Exploration"
author: "Ori Shoham"
date: "2024-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
rm(list = ls())
gc()
library(tidyverse) 
library(data.table)
library(haven)
library(glue)
library(tidymodels)
library(vip)
```

```{r}
if(str_detect(getwd(), "/code")){
  setwd(str_remove(getwd(), "/code"))
}
FYC <- fread("datasets/base_samp_2008_2010.csv")
```
The sample I start with includes panels 13-15 (beginning in 2008-2010). I only include people for whom there is data in both panel years. I am ignoring the family structure of the data for the time being, this means I will not use the survey administrative variables as predictors to avoid leakage of data within the family from train to test sets. One problem I don't currently address is that for some administrative reasons expenditure variables may represent only a subset of the year for few respondents, handling this will require diving deeper into MEPS sample definitions which seems a bit much for now.

## Expenditure to logs

```{r}
FYC <- FYC %>% 
  mutate(across(c(starts_with("exp"), starts_with("y_exp")),
                .fns = list(log = \(x) log(x+1))))
```


## Split the Data

```{r}
set.seed(2024)
FYC_split <- initial_split(FYC, prop = 0.5)
FYC_tr <- training(FYC_split)
FYC_test <- testing(FYC_split)
tr_ids <- FYC_tr$surv_ad_DUPERSID
test_ids <- FYC_test$surv_ad_DUPERSID
saveRDS(tr_ids, "interim_objects/sample_split/tr_ids.RDS")
saveRDS(test_ids, "interim_objects/sample_split/test_ids.RDS")
```



## Explore main outcome

To get things started I focus on the variable `y_exp_TOTEXP_y2` and it's log version which is the total expenditure in the second year. 

```{r}
summary(FYC_tr$y_exp_TOTEXP_y2)
# proportion 0 expenditure
mean(FYC_tr$y_exp_TOTEXP_y2 == 0)

summary(FYC_tr$y_exp_TOTEXP_y2_log)

# plot distribution of log total expenditure
FYC_tr %>% 
  ggplot(aes(x = y_exp_TOTEXP_y2_log)) + 
  geom_density() +
  theme_minimal() +
  labs(x = "log(total expenditure second year + 1)", title = "Total Expenditure Distribution")
```

## A few predictive benchmarks

```{r}

lm_rsq <- function(df, y_var, x_var){
    form <- as.formula(glue("{y_var}~{x_var}"))
    m <- lm(form, df);
    eq <- substitute(italic(RMSE) == rmse*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3),
             rmse = format(sqrt(mean(m$residuals^2)), digits = 3)))
    as.character(as.expression(eq));
}

    

# expenditure and age
FYC_tr %>% 
  ggplot(aes(x = dem_AGELAST,y = y_exp_TOTEXP_y2_log)) +
  geom_point(alpha = 0.2, color = "lightblue") +
  theme_minimal() +
  geom_smooth(method = "lm", color = "darkred", linewidth = 1) +
  annotate(geom = "text", x = 15, y = 10, label = lm_rsq(FYC_tr, "y_exp_TOTEXP_y2_log", "dem_AGELAST" ), parse= T) +
  labs(title = "Expenditure on age", x = "Age (first year)", y = "log(total expenditure second year + 1)")

mod_age <- lm(y_exp_TOTEXP_y2_log ~ dem_AGELAST, data = FYC_tr )

# Expenditure y2 on expenditure y1

FYC_tr %>% 
  ggplot(aes(x = exp_TOTEXP_y1_log,y = y_exp_TOTEXP_y2_log)) +
  geom_point(alpha = 0.2, color = "lightblue") +
  theme_minimal() +
  geom_smooth(method = "lm", color = "darkred", linewidth = 1) +
  annotate(geom = "text", x = 2.5, y = 10, label = lm_rsq(FYC_tr, "y_exp_TOTEXP_y2_log", "exp_TOTEXP_y1_log" ), parse= T) +
  labs(title = "Expenditure year 2 on expenditure year 1", x = "log(total expenditure first year + 1)", y = "log(total expenditure second year + 1)")

mod_exp <- lm(y_exp_TOTEXP_y2_log ~ exp_TOTEXP_y1_log, data = FYC_tr )

```



Linear regression benchmarks based on age and year 1 expenditure


```{r}
mod_age_lin <- lm(y_exp_TOTEXP_y2_log ~ dem_AGELAST, data = FYC_tr )
mod_age_poly <- lm(y_exp_TOTEXP_y2_log ~ poly(dem_AGELAST, degree = 4), data = FYC_tr )
mod_exp_base <- lm(y_exp_TOTEXP_y2_log ~ exp_TOTEXP_y1_log, data = FYC_tr )
mod_exp_w0s <- lm(y_exp_TOTEXP_y2_log ~ exp_TOTEXP_y1_log + I(exp_TOTEXP_y1_log == 0), data = FYC_tr )
mod_all_int <- lm(y_exp_TOTEXP_y2_log ~ exp_TOTEXP_y1_log*poly(dem_AGELAST, degree = 4) +
                  I(exp_TOTEXP_y1_log == 0)*poly(dem_AGELAST, degree = 5), data = FYC_tr )

# create data with predictions both on train and test sets from all specifications to calculate performance
tr_age_lin <- FYC_tr %>% 
    mutate(data = "train", spec = "Age linear", .pred = predict(mod_age_lin))
te_age_lin <- FYC_test %>% 
    mutate(data = "test", spec = "Age linear", .pred = predict(mod_age_lin, newdata = FYC_test))
tr_age_poly <- FYC_tr %>%
    mutate(data = "train", spec = "Age polynomial", .pred = predict(mod_age_poly))
te_age_poly <- FYC_test %>%
    mutate(data = "test", spec = "Age polynomial", .pred = predict(mod_age_poly, newdata = FYC_test))
tr_exp_base <- FYC_tr %>% 
    mutate(data = "train", spec = "Expenditure y1", .pred = predict(mod_exp_base))
te_exp_base <- FYC_test %>% 
    mutate(data = "test", spec = "Expenditure y1", .pred = predict(mod_exp_base, newdata = FYC_test))
tr_exp_w0s <- FYC_tr %>% 
    mutate(data = "train", spec = "Expenditure y1 w 0s", .pred = predict(mod_exp_w0s))
te_exp_w0s <- FYC_test %>% 
    mutate(data = "test", spec = "Expenditure y1 w 0s", .pred = predict(mod_exp_w0s, newdata = FYC_test))
tr_all_int <- FYC_tr %>%
    mutate(data = "train", spec = "Age polynomial interacted w expenditure y1", .pred = predict(mod_all_int))
te_all_int <- FYC_test %>%
    mutate(data = "test", spec = "Age polynomial interacted w expenditure y1", .pred = predict(mod_all_int, newdata = FYC_test))

# combine into one dataset
res <- bind_rows(mget(ls()[str_detect(ls(), "tr_|te_") & ls()!= "tr_ids"]))
res_sum <- res %>% 
  group_by(spec, data) %>% 
  rmse(truth = y_exp_TOTEXP_y2_log, estimate = .pred) %>% 
  bind_rows(res %>% 
    group_by(spec, data) %>% 
    rsq(truth = y_exp_TOTEXP_y2_log, estimate = .pred)) %>% 
  pivot_wider(names_from = c(data, .metric), values_from = .estimate) %>% 
  select(!.estimator) 

res_sum %>% 
  select(spec, train_rmse, test_rmse, train_rsq, test_rsq)
```


## Random forest

Start with random forest with all variables except for survey_ad and one without expenditure as well. No feature engineering at all, for now dropping variables with explicit NAs (those are the variables that are not in all three datasets)


```{r}
set.seed(2023)
# use all predictors 
rec_all <- recipe(y_exp_TOTEXP_y2_log ~., data = FYC_tr) %>% 
  update_role(starts_with("surv_ad_") & !"surv_ad_PANEL",
              starts_with("y_") & !"y_exp_TOTEXP_y2_log" , contains("_r3"),
              new_role = "discarded") %>% 
  step_rm(where(\(x) sum(is.na(x))>0))
# use all predictors except for expenditure in year 1

rec_no_exp <- recipe(y_exp_TOTEXP_y2_log ~., data = FYC_tr) %>% 
  update_role(starts_with("surv_ad_") & !"surv_ad_PANEL", starts_with("exp_"),
              starts_with("y_") & !"y_exp_TOTEXP_y2_log" , contains("_r3"),
              new_role = "discarded")%>% 
  step_rm(where(\(x) sum(is.na(x))>0))


mod_rf <- rand_forest(mode = "regression") %>% 
  set_engine("randomForest")

rec_list <- list(rec_all, rec_no_exp)

workflow_all <- workflow() %>% 
  add_recipe(rec_all) %>% 
  add_model(mod_rf)

workflow_no_exp <- workflow() %>% 
  add_recipe(rec_no_exp) %>% 
  add_model(mod_rf)

```

```{r, eval = FALSE }

fit_all <- fit(workflow_all, FYC_tr)
fit_no_exp <- fit(workflow_no_exp, FYC_tr)

saveRDS(fit_all, "interim_objects/models/rf_all_pred_13_15_unproc.RDS")
saveRDS(fit_no_exp, "interim_objects/models/rf_no_exp_13_15_unproc.RDS")
```


```{r}
if(str_detect(getwd(), "/code")){
  setwd(str_remove(getwd(), "/code"))
}
fit_all <- readRDS("interim_objects/models/rf_all_pred_13_15_unproc.RDS")
fit_no_exp <- readRDS("interim_objects/models/rf_no_exp_13_15_unproc.RDS")

results <- fit_all %>% 
  predict(FYC_tr) %>% 
  mutate(truth = FYC_tr$y_exp_TOTEXP_y2_log, predictors = "all", data = "train") %>% 
  bind_rows(fit_all %>% 
    predict(FYC_test) %>% 
    mutate(truth = FYC_test$y_exp_TOTEXP_y2_log, predictors = "all", data = "test")) %>%
  bind_rows(fit_no_exp %>% 
    predict(FYC_tr) %>% 
    mutate(truth = FYC_tr$y_exp_TOTEXP_y2_log, predictors = "no_exp", data = "train")) %>%
  bind_rows(fit_no_exp %>% 
    predict(FYC_test) %>% 
    mutate(truth = FYC_test$y_exp_TOTEXP_y2_log, predictors = "no_exp", data = "test")) %>% 
  bind_rows(mod_age %>% 
              predict(FYC_tr) %>% 
              tibble( ) %>% 
              rename(.pred = ".") %>% 
              mutate(truth = FYC_tr$y_exp_TOTEXP_y2_log, predictors = "age_lm", data = "train")) %>% 
  bind_rows(mod_age %>% 
              predict(FYC_test) %>% 
              tibble( ) %>% 
              rename(.pred = ".") %>% 
              mutate(truth = FYC_test$y_exp_TOTEXP_y2_log, predictors = "age_lm", data = "test")) %>% 
  bind_rows(mod_exp %>% 
              predict(FYC_tr) %>% 
              tibble( ) %>% 
              rename(.pred = ".") %>% 
              mutate(truth = FYC_tr$y_exp_TOTEXP_y2_log, predictors = "exp_lm", data = "train")) %>% 
  bind_rows(mod_exp %>% 
              predict(FYC_test) %>% 
              tibble( ) %>% 
              rename(.pred = ".") %>% 
              mutate(truth = FYC_test$y_exp_TOTEXP_y2_log, predictors = "exp_lm", data = "test"))

res_sum <- results %>% 
  group_by(predictors, data) %>% 
  rmse(truth = truth, estimate = .pred) %>% 
  bind_rows(results %>% 
    group_by(predictors, data) %>% 
    rsq(truth = truth, estimate = .pred)) %>% 
  pivot_wider(names_from = c(data, .metric), values_from = .estimate) %>% 
  select(!.estimator) 

res_sum[c(1,3,2,4),] %>% 
  select(predictors, train_rmse, test_rmse, train_rsq, test_rsq)
# variable importance all predictors
imp_all <- fit_all %>% 
  vip(num_features = 25)

orig_var_names <- imp_all$data$Variable

imp_all$data$Variable <- 
  c("Tot. Exp. y_1", "Log Tot. Exp. y_1", "# Presc. Medicines (incl. refills) y_1",
    "Log To. Exp. on Presc. Meds (incl. refills) y_1", "Tot. Exp. on Presc. Meds (incl. refills) y_1",
    "# Office-Based Provider Visits y_1", "Log Tot. Charegs (excl. presc. meds) y_1", 
    "Tot. Charegs (excl. presc. meds) y_1", "Office-Based Provider Visits Charges y_1", "Variance Estimation Stratum",
    "Family Inc as % Poverty Line", "Log Office-Based Provider Visits Charges y_1", "FINAL PERSON WEIGHT y_1", 
    "Family's Total Income", "POV ADJ FAMILY WGT-CPS FAM", "FINAL FAMILY WEIGHT", "Tot. OOP Exp. on Presc. Meds y_1",
    "Tot. OFFICE-BASED EXP. y_1", "Log Tot. OFFICE-BASED EXP. y_1",  "Log Tot. OOP Exp. on Presc. Meds y_1", "Birth Year",
    "Birth Month", "Imputed Age round 2", "SAQ:PHY COMPONENT SUMMRY round 2", "SAQ:MNT COMPONENT SUMMRY round 2")
imp_all +
  theme_classic()


# variable importance no expenditure
imp_no_exp <- fit_no_exp %>% 
  vip(num_features = 25)

for(x in imp_no_exp$data$Variable ){
  if (x %in% orig_var_names){
    imp_no_exp$data$Variable[imp_no_exp$data$Variable == x] <- imp_all$data$Variable[orig_var_names == x]
  }
}
imp_no_exp$data$Variable[c(3,4,5,6,13,14,16,18,19,20,21,23,24) ] <- c(
  "# OFFICE-BASED PHYSICIAN VISITS y_1", "HAS USUAL 3RD PARTY PAYER FOR PMEDS round 2",
  "HAS USUAL 3RD PARTY PAYER FOR PMEDS round 1", "# DENTAL CARE VISITS y_1", "ALL OFFICE VISITS -PRV & TRI AMT",
  "INSURED end y_1", "SAQ 12 MOS: # VISITS TO MED OFF FOR CARE round 2",
  "Age end of ref period", "Age y_1 imputed", "# GENERAL DENTIST VISITS y_1", "COVR BY HOSP/MED INS IN end y_1",
  "OTAL presc med -PRV & TRI AMT y_1", "Uninsured all y_1"
)

imp_no_exp + 
  theme_classic()
```

